---
title: "Getting Started with UWBiost561"
author: "Kevin Simiyu"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_float:
      toc_collapsed: true
    theme: united
vignette: >
  %\VignetteIndexEntry{Getting Started with UWBiost561}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette demonstrates the core functionality of the `UWBiost561` package.
The package provides tools for:

1. Generating random adjacency matrices with embedded partial cliques
2. Computing the maximal partial clique from an adjacency matrix

No external data is needed — all examples use synthetic data generated
directly by the package functions.

## Generating a Random Graph with a Partial Clique

The `generate_partial_clique()` function creates a random adjacency matrix
with an embedded partial clique of a specified size and edge density.

```{r}
library(UWBiost561)

set.seed(10)
simulation <- UWBiost561::generate_partial_clique(
  n = 10,
  clique_fraction = 0.5,
  clique_edge_density = 0.9
)
```

The function returns a list. The main output is `adj_mat`, a symmetric
binary matrix where `1` indicates an edge between two nodes:

```{r}
simulation$adj_mat
```

Here, `n = 10` means 10 nodes, `clique_fraction = 0.5` means 5 of those
nodes form the partial clique, and `clique_edge_density = 0.9` means 90%
of the possible edges within the clique are present.

## Visualizing the Adjacency Matrix

We can visualize the adjacency matrix as a heatmap to see the structure
of the graph:

```{r, fig.align="center", fig.width=5, fig.height=5}
image(simulation$adj_mat,
      main = "Adjacency Matrix Heatmap",
      col = c("white", "steelblue"),
      xaxt = "n", yaxt = "n")
```

The darker region in the upper-left corner corresponds to the embedded
partial clique among the first few nodes.

## Computing the Maximal Partial Clique

The `compute_maximal_partial_clique()` function takes an adjacency matrix
and a required edge density `alpha`, and returns the largest subset of
nodes that forms a partial clique at that density.

```{r}
res <- UWBiost561::compute_maximal_partial_clique(
  adj_mat = simulation$adj_mat,
  alpha = 0.9
)

res
```

- `clique_idx`: the node indices identified as part of the maximal partial clique
- `edge_density`: the actual edge density among those nodes (should be >= `alpha`)

## Verifying the Result

We can verify that the returned nodes satisfy the required edge density:

```{r}
idx <- res$clique_idx
m <- length(idx)
submat <- simulation$adj_mat[idx, idx]

# Number of edges (excluding diagonal)
num_edges <- (sum(submat) - m) / 2
max_edges <- m * (m - 1) / 2

cat("Nodes in clique:  ", idx, "\n")
cat("Clique size:      ", m, "\n")
cat("Edges found:      ", num_edges, "\n")
cat("Max possible:     ", max_edges, "\n")
cat("Edge density:     ", round(num_edges / max_edges, 3), "\n")
cat("Required alpha:    0.9\n")
```

## Scaling Up

Let's try a larger graph to see how the functions scale:

```{r}
set.seed(42)
simulation_large <- UWBiost561::generate_partial_clique(
  n = 20,
  clique_fraction = 0.4,
  clique_edge_density = 0.85
)

res_large <- UWBiost561::compute_maximal_partial_clique(
  adj_mat = simulation_large$adj_mat,
  alpha = 0.85
)

cat("Graph size:        20 nodes\n")
cat("Expected clique:  ", round(20 * 0.4), "nodes\n")
cat("Found clique:     ", length(res_large$clique_idx), "nodes\n")
cat("Edge density:     ", round(res_large$edge_density, 3), "\n")
```

## Summary

The `UWBiost561` package provides a simple workflow for studying partial
cliques in random graphs:

1. Use `generate_partial_clique()` to simulate a graph with a known partial clique
2. Use `compute_maximal_partial_clique()` to recover the partial clique
3. Verify the result by checking the edge density of the returned nodes
